\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{parskip}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}

\setlength{\emergencystretch}{2em}

\begin{document}
\thispagestyle{empty}

\noindent\textbf{Phuc Hao Do}\\
Faculty of Information Technology\\
Danang Architecture University, Danang, Vietnam\\
\href{mailto:haodp@dau.edu.vn}{haodp@dau.edu.vn}

\vspace{1.2em}
\noindent June 27, 2026

\vspace{1.2em}
\noindent\textbf{To:} The Guest Editors,\\
IEEE Transactions on Network Science and Engineering,\\
Special Issue on \emph{Tutorials and Surveys on AI and Quantum-enabled Learning for
Future Wireless Networks}\\
(Prof. Trung Q. Duong, Prof. Ruidong Li, Dr. Bhaskara Narottama, Prof. George K.
Karagiannidis)

\vspace{1.0em}
\noindent\textbf{Re:} Submission of the manuscript ``\emph{Quantum and Quantum-Inspired
Graph Neural Networks for Dynamic Wireless Network Topologies: A Tutorial and Survey with
a Reproducible LEO Case Study}''

\vspace{1.2em}
\noindent Dear Guest Editors,

I am pleased to submit the above manuscript for consideration in the Special Issue on
\emph{Tutorials and Surveys on AI and Quantum-enabled Learning for Future Wireless
Networks}. The paper is prepared as a combined \textbf{tutorial and survey}, the
contribution type the call solicits, and it addresses several of the listed topics
directly, including \emph{quantum graph neural networks}, \emph{data-driven quantum
algorithms for dynamic topologies}, \emph{near-term quantum computing for routing in
non-terrestrial networks}, and \emph{non-centralized learning of classical and quantum
data}.

\textbf{What the paper offers.} The manuscript is organized around a single unifying idea:
classical graph neural networks, graph diffusion, and continuous-time quantum walks are
the same object, a propagation operator on the graph Laplacian, so the only thing that
changes between them is the spreading physics (diffusive for the heat kernel, ballistic
for the quantum walk). The method is therefore explicit and simple to state: write every
model as a propagation operator acting on graph signals, vary only that operator on a
parameter-matched backbone, and measure honestly. From this view the paper (i) surveys
the classical, quantum, quantum-inspired, and dynamic-graph branches and their use in
wireless and non-terrestrial networks, organized by operator rather than by author; (ii)
teaches, with pseudocode, a formal result (a proposition deriving the diffusive-versus-ballistic
spreading laws exactly on the line graph), and a worked example, how to build a
parameter-matched quantum-inspired GNN layer that runs on classical hardware today; and
(iii) demonstrates the framework on a fully reproducible low Earth orbit (LEO) case study.

\textbf{The central finding is honest and, I believe, timely for this Special Issue.} On a
node-potential task over a dynamic LEO graph, global propagation beats local message
passing by a wide margin at every scale, and the quantum-inspired and classical operators
trade places as the constellation grows: the quantum-inspired walk loses at 66 satellites,
ties at 264, and becomes the best operator at 1584 satellites, exactly the
diameter-driven trend its ballistic spreading predicts. The paper reports this
conditional, scale-dependent quantum advantage with its caveats rather than as a blanket
claim, and adds a NISQ-noise study (PennyLane) showing how device noise erodes the
mechanism. A dedicated section on evaluation pitfalls, scale-coupled normalization and
single-seed reporting, distills a reusable checklist that should be useful to the
community independently of the case study.

\textbf{Reproducibility.} Every figure and number in the paper is regenerated from seeded
runs by the companion code at
\href{https://github.com/haodpsut/qi-gnn-dynamic-wireless}{github.com/haodpsut/qi-gnn-dynamic-wireless}.
All experiments are parameter-matched and multi-seed.

\textbf{Originality.} This manuscript is original, has not been published before, and is
not under consideration elsewhere. The LEO topology generator reuses my open-source
simulator, while the node-potential task, the experiments, and all results here are new.

I confirm there are no conflicts of interest with the Guest Editors that would prevent a
fair review, and I am happy to suggest or exclude reviewers if helpful.

Thank you for your time and consideration.

\vspace{1.4em}
\noindent Sincerely,

\vspace{0.4em}
\noindent\textbf{Phuc Hao Do}\\
Danang Architecture University

\end{document}
